Combined annotation exploration for SCPCP000006

### 2025-02-05 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Update 07_combined_annotation_across_samples_exploration.Rmd to pass the cli test run

---
title: "Combined annotation exploration for SCPCP000006"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
params:
  predicted.celltype.threshold: 0.85
  cnv_threshold_low: 0
  cnv_threshold_high: 2
  testing: 0
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

        # proportion of cnv per chromosome
        proportion_cnv_chr1 = srat$proportion_cnv_chr1,
        proportion_cnv_chr2 = srat$proportion_cnv_chr2,
        proportion_cnv_chr3 = srat$proportion_cnv_chr3,
        proportion_cnv_chr4 = srat$proportion_cnv_chr4,
        proportion_cnv_chr5 = srat$proportion_cnv_chr5,
        proportion_cnv_chr6 = srat$proportion_cnv_chr6,
        proportion_cnv_chr7 = srat$proportion_cnv_chr7,
        proportion_cnv_chr8 = srat$proportion_cnv_chr8,
        proportion_cnv_chr9 = srat$proportion_cnv_chr9,
        proportion_cnv_chr10 = srat$proportion_cnv_chr10,
        proportion_cnv_chr11 = srat$proportion_cnv_chr11,
        proportion_cnv_chr12 = srat$proportion_cnv_chr12,
        proportion_cnv_chr13 = srat$proportion_cnv_chr13,
        proportion_cnv_chr14 = srat$proportion_cnv_chr14,
        proportion_cnv_chr15 = srat$proportion_cnv_chr15,
        proportion_cnv_chr16 = srat$proportion_cnv_chr16,
        proportion_cnv_chr17 = srat$proportion_cnv_chr17,
        proportion_cnv_chr18 = srat$proportion_cnv_chr18,
        proportion_cnv_chr19 = srat$proportion_cnv_chr19,
        proportion_cnv_chr20 = srat$proportion_cnv_chr20,
        proportion_cnv_chr21 = srat$proportion_cnv_chr21,
        proportion_cnv_chr22 = srat$proportion_cnv_chr22,

        # cnv global estimation per chromosome
        has_cnv_chr1 = srat$has_cnv_chr1,
        has_cnv_chr2 = srat$has_cnv_chr2,
        has_cnv_chr3 = srat$has_cnv_chr3,
        has_cnv_chr4 = srat$has_cnv_chr4,
        has_cnv_chr5 = srat$has_cnv_chr5,
        has_cnv_chr6 = srat$has_cnv_chr6,
        has_cnv_chr7 = srat$has_cnv_chr7,
        has_cnv_chr8 = srat$has_cnv_chr8,
        has_cnv_chr9 = srat$has_cnv_chr9,
        has_cnv_chr10 = srat$has_cnv_chr10,
        has_cnv_chr11 = srat$has_cnv_chr11,
        has_cnv_chr12 = srat$has_cnv_chr12,
        has_cnv_chr13 = srat$has_cnv_chr13,
        has_cnv_chr14 = srat$has_cnv_chr14,
        has_cnv_chr15 = srat$has_cnv_chr15,
        has_cnv_chr16 = srat$has_cnv_chr16,
        has_cnv_chr17 = srat$has_cnv_chr17,
        has_cnv_chr18 = srat$has_cnv_chr18,
        has_cnv_chr19 = srat$has_cnv_chr19,
        has_cnv_chr20 = srat$has_cnv_chr20,
        has_cnv_chr21 = srat$has_cnv_chr21,
        has_cnv_chr22 = srat$has_cnv_chr22
      ) |>
        tibble::rownames_to_column("barcode") |>
        dplyr::mutate(sample_id = sample_id)
    }
  ) |>
  # now combine all dataframes to make one big one
  dplyr::bind_rows()
```

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
cell_type_df <- cell_type_df |>
  mutate(has_cnv_score = rowSums(cell_type_df[, grepl("has_cnv_chr", colnames(cell_type_df))])) |>
  mutate(has_cnv_score = case_when(
    has_cnv_score > params$cnv_threshold_high ~ "CNV",
    has_cnv_score <= params$cnv_threshold_low ~ "no CNV",
    .default = "unknown"
  ))

@@

- _normal_ cells can be observe in all four compartments (`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV
We only allow a bit of flexibility in terms of CNV profile for immune and endothelium cells that have a high predicted score.
Indeed, we know that false positive CNV can be observed in a cell type specific manner.

@@

    compartment %in% c("endothelium", "immune") &
      cell_type.score > params$predicted.celltype.threshold ~ "normal",

@@

    cell_type_df$compartment %in% c("endothelium") &
      (cell_type_df$compartment.score > params$predicted.celltype.threshold) ~ "endothelium",
    cell_type_df$compartment %in% c("immune") &
      (cell_type_df$compartment.score > params$predicted.celltype.threshold) ~ "immune",

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
  geom_point(size = 0.2, shape = 19) +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
tmp <- cell_type_df[cell_type_df$second.level_annotation != "unknown",]
ggplot(tmp, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
  geom_point(size = 0.2, shape = 19) +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df[cell_type_df$first.level_annotation == "cancer", ], aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
  geom_point(size = 0.2, shape = 19) +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
for (i in 1:22) {
  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_cnv_chr", i)))
}
```

@@

```{r}
annotations_table <- cell_type_df |>
  dplyr::select(
    cell_barcode = barcode,
    scpca_sample_id = sample_id,
    tumor_cell_classification = first.level_annotation,
    cell_type_assignment = second.level_annotation
  ) |>
  mutate(
    # change cancer --> tumor, but keep the other labels
    tumor_cell_classification = ifelse(
      tumor_cell_classification == "cancer", "tumor", tumor_cell_classification
    ),
    cell_type_assignment = str_replace_all(
      cell_type_assignment,
      "cancer ",
      "tumor "
    )
  )

@@

- Combining label transfer and CNV inference we have produced draft annotations for 35/40 Wilms tumor samples in SCPCP000006.
We decided not to annotate samples for which the `infercnv` couldn't be run with a normal reference, as we don't trust the results in that case and want to avoid misclassification (`SCPCS000177`, `SCPCS000180`, `SCPCS000181`, `SCPCS000190`, `SCPCS000197`).

>>>>

---
title: "Combined annotation exploration for SCPCP000006"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
params:
  predicted.celltype.threshold: 0.75
  cnv_threshold_low: 0
  cnv_threshold_high: 2
  testing: 0
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

        # proportion of cnv per chromosome
        proportion_cnv_chr1p = srat$proportion_cnv_chr1p,
        proportion_cnv_chr2p = srat$proportion_cnv_chr2p,
        proportion_cnv_chr3p = srat$proportion_cnv_chr3p,
        proportion_cnv_chr4p = srat$proportion_cnv_chr4p,
        proportion_cnv_chr5p = srat$proportion_cnv_chr5p,
        proportion_cnv_chr6p = srat$proportion_cnv_chr6p,
        proportion_cnv_chr7p = srat$proportion_cnv_chr7p,
        proportion_cnv_chr8p = srat$proportion_cnv_chr8p,
        proportion_cnv_chr9p = srat$proportion_cnv_chr9p,
        proportion_cnv_chr10p = srat$proportion_cnv_chr10p,
        proportion_cnv_chr11p = srat$proportion_cnv_chr11p,
        proportion_cnv_chr12p = srat$proportion_cnv_chr12p,

        proportion_cnv_chr16p = srat$proportion_cnv_chr16p,
        proportion_cnv_chr17p = srat$proportion_cnv_chr17p,
        proportion_cnv_chr18p = srat$proportion_cnv_chr18p,
        proportion_cnv_chr19p = srat$proportion_cnv_chr19p,
        proportion_cnv_chr20p = srat$proportion_cnv_chr20p,

        proportion_cnv_chr1q = srat$proportion_cnv_chr1q,
        proportion_cnv_chr2q = srat$proportion_cnv_chr2q,
        proportion_cnv_chr3q = srat$proportion_cnv_chr3q,
        proportion_cnv_chr4q = srat$proportion_cnv_chr4q,
        proportion_cnv_chr5q = srat$proportion_cnv_chr5q,
        proportion_cnv_chr6q = srat$proportion_cnv_chr6q,
        proportion_cnv_chr7q = srat$proportion_cnv_chr7q,
        proportion_cnv_chr8q = srat$proportion_cnv_chr8q,
        proportion_cnv_chr9q = srat$proportion_cnv_chr9q,
        proportion_cnv_chr10q = srat$proportion_cnv_chr10q,
        proportion_cnv_chr11q = srat$proportion_cnv_chr11q,
        proportion_cnv_chr12q = srat$proportion_cnv_chr12q,
        proportion_cnv_chr13q = srat$proportion_cnv_chr13q,
        proportion_cnv_chr14q = srat$proportion_cnv_chr14q,
        proportion_cnv_chr15q = srat$proportion_cnv_chr15q,
        proportion_cnv_chr16q = srat$proportion_cnv_chr16q,
        proportion_cnv_chr17q = srat$proportion_cnv_chr17q,
        proportion_cnv_chr18q = srat$proportion_cnv_chr18q,
        proportion_cnv_chr19q = srat$proportion_cnv_chr19q,
        proportion_cnv_chr20q = srat$proportion_cnv_chr20q,
        proportion_cnv_chr21q = srat$proportion_cnv_chr21q,
        proportion_cnv_chr22q = srat$proportion_cnv_chr22q,

        # loss global estimation per chromosome
        has_loss_chr1p = srat$has_loss_chr1p,
        has_loss_chr2p = srat$has_loss_chr2p,
        has_loss_chr3p = srat$has_loss_chr3p,
        has_loss_chr4p = srat$has_loss_chr4p,
        has_loss_chr5p = srat$has_loss_chr5p,
        has_loss_chr6p = srat$has_loss_chr6p,
        has_loss_chr7p = srat$has_loss_chr7p,
        has_loss_chr8p = srat$has_loss_chr8p,
        has_loss_chr9p = srat$has_loss_chr9p,
        has_loss_chr10p = srat$has_loss_chr10p,
        has_loss_chr11p = srat$has_loss_chr11p,
        has_loss_chr12p = srat$has_loss_chr12p,

        has_loss_chr16p = srat$has_loss_chr16p,
        has_loss_chr17p = srat$has_loss_chr17p,
        has_loss_chr18p = srat$has_loss_chr18p,
        has_loss_chr19p = srat$has_loss_chr19p,
        has_loss_chr20p = srat$has_loss_chr20p,

        has_loss_chr1q = srat$has_loss_chr1q,
        has_loss_chr2q = srat$has_loss_chr2q,
        has_loss_chr3q = srat$has_loss_chr3q,
        has_loss_chr4q = srat$has_loss_chr4q,
        has_loss_chr5q = srat$has_loss_chr5q,
        has_loss_chr6q = srat$has_loss_chr6q,
        has_loss_chr7q = srat$has_loss_chr7q,
        has_loss_chr8q = srat$has_loss_chr8q,
        has_loss_chr9q = srat$has_loss_chr9q,
        has_loss_chr10q = srat$has_loss_chr10q,
        has_loss_chr11q = srat$has_loss_chr11q,
        has_loss_chr12q = srat$has_loss_chr12q,
        has_loss_chr13q = srat$has_loss_chr13q,
        has_loss_chr14q = srat$has_loss_chr14q,
        has_loss_chr15q = srat$has_loss_chr15q,
        has_loss_chr16q = srat$has_loss_chr16q,
        has_loss_chr17q = srat$has_loss_chr17q,
        has_loss_chr18q = srat$has_loss_chr18q,
        has_loss_chr19q = srat$has_loss_chr19q,
        has_loss_chr20q = srat$has_loss_chr20q,
        has_loss_chr21q = srat$has_loss_chr21q,
        has_loss_chr22q = srat$has_loss_chr22q,

        # dupli global estimation per chromosome
        has_dupli_chr1p = srat$has_dupli_chr1p,
        has_dupli_chr2p = srat$has_dupli_chr2p,
        has_dupli_chr3p = srat$has_dupli_chr3p,
        has_dupli_chr4p = srat$has_dupli_chr4p,
        has_dupli_chr5p = srat$has_dupli_chr5p,
        has_dupli_chr6p = srat$has_dupli_chr6p,
        has_dupli_chr7p = srat$has_dupli_chr7p,
        has_dupli_chr8p = srat$has_dupli_chr8p,
        has_dupli_chr9p = srat$has_dupli_chr9p,
        has_dupli_chr10p = srat$has_dupli_chr10p,
        has_dupli_chr11p = srat$has_dupli_chr11p,
        has_dupli_chr12p = srat$has_dupli_chr12p,

        has_dupli_chr16p = srat$has_dupli_chr16p,
        has_dupli_chr17p = srat$has_dupli_chr17p,
        has_dupli_chr18p = srat$has_dupli_chr18p,
        has_dupli_chr19p = srat$has_dupli_chr19p,
        has_dupli_chr20p = srat$has_dupli_chr20p,

        has_dupli_chr1q = srat$has_dupli_chr1q,
        has_dupli_chr2q = srat$has_dupli_chr2q,
        has_dupli_chr3q = srat$has_dupli_chr3q,
        has_dupli_chr4q = srat$has_dupli_chr4q,
        has_dupli_chr5q = srat$has_dupli_chr5q,
        has_dupli_chr6q = srat$has_dupli_chr6q,
        has_dupli_chr7q = srat$has_dupli_chr7q,
        has_dupli_chr8q = srat$has_dupli_chr8q,
        has_dupli_chr9q = srat$has_dupli_chr9q,
        has_dupli_chr10q = srat$has_dupli_chr10q,
        has_dupli_chr11q = srat$has_dupli_chr11q,
        has_dupli_chr12q = srat$has_dupli_chr12q,
        has_dupli_chr13q = srat$has_dupli_chr13q,
        has_dupli_chr14q = srat$has_dupli_chr14q,
        has_dupli_chr15q = srat$has_dupli_chr15q,
        has_dupli_chr16q = srat$has_dupli_chr16q,
        has_dupli_chr17q = srat$has_dupli_chr17q,
        has_dupli_chr18q = srat$has_dupli_chr18q,
        has_dupli_chr19q = srat$has_dupli_chr19q,
        has_dupli_chr20q = srat$has_dupli_chr20q,
        has_dupli_chr21q = srat$has_dupli_chr21q,
        has_dupli_chr22q = srat$has_dupli_chr22q
      ) |>
        tibble::rownames_to_column("barcode") |>
        dplyr::mutate(sample_id = sample_id)
    }
  ) |>
  # now combine all dataframes to make one big one
  dplyr::bind_rows()
```

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
cell_type_df <- cell_type_df |>
  mutate(has_cnv_score = rowSums(cell_type_df[, grepl("has_loss_chr|has_dupli_chr", colnames(cell_type_df))])) |>
  mutate(has_cnv_score = case_when(
    has_cnv_score > params$cnv_threshold_high ~ "CNV",
    has_cnv_score <= params$cnv_threshold_low ~ "no CNV",
    .default = "unknown"
  ))

@@

- _normal_ cells can be observe in all five compartments (`normal`,`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV
We only allow a bit of flexibility in terms of CNV profile for immune and endothelium cells that have a high predicted score.
Please note that in `06_infercnv.R`, we renamed all `immune` and `endothelium` cells with a `fetal_kidney_predicted.compartment.score` > `r params$predicted.celltype.threshold` as `normal`.
Indeed, we know that false positive CNV can be observed in a cell type specific manner.

@@

    compartment %in% c("normal", "immune", "endothelium") &
      cell_type.score > params$predicted.celltype.threshold ~ "normal",

@@

    cell_type_df$compartment %in% c("normal") &
      cell_type_df$cell_type %in% c("endothelial cell") ~ "endothelium",
    cell_type_df$compartment %in% c("normal") &
      cell_type_df$cell_type %in% c("B cell",
                                    "CD4-positive, alpha-beta T cell",
                                    "conventional dendritic cell",
                                    "lymphocyte",
                                    "macrophage",
                                    "mast cell",
                                    "monocyte",
                                    "natural killer cell",
                                    "neutrophil",
                                    "plasmacytoid dendritic cell") ~ "immune",

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
  geom_point(size = 1, shape = 19) +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
tmp <- cell_type_df[cell_type_df$second.level_annotation != "unknown",]
ggplot(tmp, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
  geom_point(size = 1, shape = 19) +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df[cell_type_df$first.level_annotation == "cancer", ], aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
  geom_point(size = 1, shape = 19) +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

@@

#### Loss on long arms

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
for (i in 1:22) {
  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "q")))
}
```
#### gains on long arms

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
for (i in 1:22) {
  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "q")))
}
```

#### loss on short arms

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
for (i in c(1:12, 16:20)) {
  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "p")))
}
```

#### gains on short arms

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
for (i in c(1:12, 16:20)) {
  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "p")))
}
```

@@

```{r}
annotations_table <- cell_type_df |>
  select(
    cell_barcode = barcode,
    scpca_sample_id = sample_id,
    tumor_cell_classification = first.level_annotation,
    cell_type_assignment = second.level_annotation
  ) |>
  mutate(
    # change cancer --> tumor, but keep the other labels
    tumor_cell_classification = ifelse(
      tumor_cell_classification == "cancer", "tumor", tumor_cell_classification
    ),
    cell_type_assignment = str_replace_all(
      cell_type_assignment,
      "cancer ",
      "tumor "
    )
  )

@@

- Combining label transfer and CNV inference we have produced draft annotations for all 40 Wilms tumor samples in SCPCP000006
